\textsc{RILS}-\textsc{ROLS} : Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-11-14 paper_resources/doc/paper.tex
Finished revision of abstract and first section.

	% first the title is needed
	\title{\textsc{Rils}-\textsc{Rols} : Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares}

	\author[1]{Aleksandar Kartelj}
	\author[2]{Marko Djukanovi\'c}
	\address[1]{$kartelj@matf.bg.ac.rs$, \\  Faculty of Mathematics, University of Belgrade, Serbia}
	\address[2]{$ marko.djukanovic@pmf.unibl.org$,\\   Faculty of Natural Sciences and Mathematics, University of Banja Luka, Bosnia and Herzegovina}
	\begin{abstract}
		In this paper, we solve the well-known, prominent  Symbolic regression problem, intensively studied in the literature within the last two decades. This problem has a wide range of applications;  for example, the obtained SR models could serve theoretically, providing insight towards establishing physical theory behind, thus helping to establish reasonable hypotheses.  Practical applications include revealing complex ecological dynamics in ecology, discovering mutations effects on protein stability in biology, modelling analytic representations of the exciton binding energy in energy science,   finding models for band gaps of NaCl-type compounds
		in material science, just to name a few. We propose a new meta-heuristic-based approach, called \textsc{Rils}-\textsc{Rols} algorithm to solve this problem. There are several key concepts utilized in this algorithm: ($i$) Iterated local search scheme is used as the backbone; ($ii$) ordinary least square method is used in the search to make it efficient by quickly determining the best--fitting   coefficients in considered solutions (models);  ($iii$) a novel fitness function is utilized which combines the three known measures: $R^2$ score, RMSE score, and the size of the model which is weighted by an internal parameter. The experiments are conducted on the two ground-truth benchmark sets, \textsc{Feynman} and \textsc{Strogatz} comparing \textsc{Rils}-\textsc{Rols} algorithm to 14 other competitors from the literature with the presence of different levels of the Gaussian white noise in the input data.  Our method is able to obtain new best exact percentages on both benchmark sets compared to other literature approaches regardless of varying the level of nose. Moreover, the performance of \textsc{Rils}-\textsc{Rols} is less sensitive with the increase of the level of noise in the data, as it is the other approaches. 
		Additionally, a new unbiased randomly generated set of problem instances are introduced where the scalability of our method is thoroughly demonstrated in terms of various instance properties. \textsc{Rils}-\textsc{Rols} method can be considered as a new state-of-the-art method for solving the problem of Symbolic regression. %for the non blck box problem instances?
	\end{abstract}
	\maketitle

@@

	The problem of symbolic regression (SR)~\cite{billard2002symbolic} has attracted many researchers over the last decade to study it intensively. SR can be seen as a generalization of the well-known concept of linear regression, i.e., polynomial regression~\cite{stimson1978interpreting}. All regression models in principle have the same task: given a set of $n$-dimensional input data and the output data, the aim is to find a  mathematical expression (function) consisting of $n$ (input) variables that fit the output data w.r.t. some in advance known measure.  This computationally intensive task is in general provenly NP--hard~\cite{virgolin2022symbolic}. When aiming a model to be a linear combination of input variables, it represents the problem of linear regression. However, when there are some nonlinear relations between variables, which are relevant, linear regression models are not enough. This is the point where symbolic regression comes into play. Unlike linear regression, it allows the search over the space of all possible mathematical formulas to find the best-fitting ones able to predict the output variable from the input variables. The basis of constructing the explicit formula is the basic operations like addition and multiplication, as well as polynomial, trigonometric, exponential, and other elementary functions.  
	%Application: https://towardsdatascience.com/real-world-applications-of-symbolic-regression-2025d17b88ef

	Concrete examples of the application of SR include obtaining a less black--box tool, as it may say more about how SR models achieve predictions, that is the coefficients and functions that build the model may indicate on larger importance of some variables over the others. Moreover, in that way, we may grasp why the variables are related in the obtained way. As an example, the appearance of an exponential function may associate with some physical phenomenon such as the intensity of radiation or acceleration over time, see~\cite{udrescu2020ai}. Additionally, the SR models, if they are correct, may possess larger extrapolation power than other mathematical models, especially better from the neural network--based methods. SR models may maximize obtaining physically more reasonable equations that could serve as an insight towards establishing physical theory behind as the final goal.  Practical applications of SR in chemical and biological sciences are listed in~\cite{weng2020simple}. In particular,  the discovery of a series of new oxide perovskite catalysts with improved activities can be found there. The application of SR to discovering physical laws from a distorted video is studied in~\cite{udrescu2021symbolic} by means of an unsupervised learning method. Revealing complex ecological dynamics by SR is presented in~\cite{chen2019revealing}, tackled by a machine learning method. Application of SR to model the effects of the mutations on protein stability, in the domain of fundamental and applied biology,  is shown in~\cite{louis2021reviewing}. One of the recent work studies of auto-discovering conserved quantities using trajectory data from unknown dynamical systems, where SR is solved by a machine learning algorithm, see~\cite{liu2021machine}. Last but not least, we mention the paper~\cite{liang2019phillips} which discovers the application of SR to model analytic representations of the exciton binding energy. The use of solving the SR in material science is described in~\cite{wang2019symbolic,wang2022symbolic,burlacu2022symbolic,kabliman2021application}. Yet another application to wind speed forecasting is given in~\cite{abdellaoui2021symbolic}.

	There are many different ways to tackle the SR by achieving the right analytic expressions, most of them are related to machine learning techniques or the technique of genetic programming (GP) and various approximate methods, such as meta-heuristics. Among the first GP methods to tackle SR is the one of Raidl~\cite{raidl1998hybrid} based on a hybrid variant of genetic programming. A differential evolution algorithm was proposed by Cerny et al.~\cite{cerny2008using}. Age-fitness Pareto Optimization approach is proposed by Smidt and Lipson~\cite{schmidt2010age}.  Application of the artificial bee colony programming to solve SR is proposed by Karaboga et al.~\cite{karaboga2012artificial}. 
	The influence of local--based heuristics on solving SR is reported by Commenda in his PhD thesis~\cite{kommenda2018local}. A  GP-based approach, the gene-pool optimal mixing evolutionary algorithm (\textsc{Gomea}) is studied by Virgolin et al.~\cite{virgolin2021improving}.  Another evolutionary algorithm, the interaction-transformation EA (\textsc{Itea}) has been proposed by de Franca et al.~\cite{de2021interaction}. Simulated annealing to solve SR is proposed by Kantor~\cite{kantor2021simulated}. A Variable neighbourhood programming approach to solve SR  is proposed by Elleurich et al.~\cite{elleuch2020variable}; this technique is initially proposed in~\cite{elleuch2016variable}. 
	Kommenda et al.~\cite{kommenda2020parameter} proposed a method called \textsc{Operon} algorithm, which uses nonlinear least squares for parameter identification of SR models further integrated into a local search mechanism in tree-based GP. The C++ implementation of \textsc{Operon} is discussed in~\cite{burlacu2020operon}. The method that utilizes the Taylor polynomials to approximate the symbolic equation that fits the dataset, called Taylor genetic programming, is proposed in~\cite{he2022taylor}. A GP approach that uses the idea of semantic back-propagation (Sbp-Gp) is proposed in~\cite{virgolin2019linear}.   Empirical analysis of variance between many GP--based methods for SR is discussed in~\cite{kammerer2021empirical}. It is also worth mentioning the GP--based Eurequa commercial solver~\cite{schmidt2009distilling, schmidt2011machine} that uses age-fitness pareto optimization with co-evolved fitness estimation. This solver is nowadays accepted as the gold standard of symbolic regression.

	Machine learning--based method based on Bayesian symbolic regression (\textsc{Bsr}) is proposed in~\cite{jin2019bayesian},  belongs to the family of Markov Chain Monte Carlo algorithms. Deep Symbolic Regression (\textsc{Dsr}), an RNN approach, is proposed in~\cite{petersen2019deep} utilizing the policy gradient search. This mechanism of search is further investigated in~\cite{landajuela2021improving}. A fast neural network approach, called \textsc{OccamNet},  is proposed in~\cite{costa2020fast}.  A deep reinforcement learning approach enhanced with genetic programming is given in~\cite{mundhenk2021symbolic}.

    Powerful hybrid techniques to solve SR are also well studied; among them, we emphasize the \textsc{Eplex} solver from~\cite{la2019probabilistic,la2016epsilon} and AI Feynman algorithm from~\cite{udrescu2020ai}, a physics-inspired divide-and-conquer method combined with the neural network fitting. The latter is one of the most efficient methods for physically inspired models. We also mention the Fast Function Extraction (\textsc{Ffx}) algorithm developed by McConaghy~\cite{mcconaghy2011ffx},  which is a non-evolutionary method   combined with a machine learning technique---path-wise regularized learning--- which prunes quickly a huge set of candidate basis functions down to compact models.

	A short overview of the most important literature methods to solve SR is given in Table~\ref{tab:gp-based} in the chronological order of their appearances.

	\begin{table}[!ht]
		\centering
		\scalebox{0.7}{
			\begin{tabularx}{550pt}{l  l  X}  
				Algorithm          &   Paper/year &   Short details   \\ \hline
				\textsc{Gp}        &    \cite{koza1994genetic} (1994)    & Application of GP to SR \\ 
				\textsc{Hybrid-Gp}  &   \cite{raidl1998hybrid} (1998)   & GP--based method; solutions locally optimizes by method of least squares to find optimum coefficients for top-level terms \\ 
				\textsc{Dfe}        & \cite{cerny2008using} (2008)   & Differential evolution algorithm \\
				\textsc{Apf-Fe}                & \cite{schmidt2009distilling, schmidt2011machine} (2009, 2011) & Age-fitness pareto optimization approach using co-evolved fitness estimation   \\
				APF            &      \cite{schmidt2010age} (2010)        &  Age-fitness pareto optimization approach                        \\
				\textsc{Ffx}  & \cite{mcconaghy2011ffx} (2011)  & The Fast Function Extraction algorithm --  non-evolutionary technique based on a machine learning technique called path-wise regularized learning \\
				\textsc{Abcp} &  \cite{karaboga2012artificial} (2012)  & Artificial bee colony programming  approach \\
				\textsc{Eplex}          &      \cite{la2016epsilon} (2016)         &   A parent selection method called $\epsilon$--lexicase selection  \\ 
				\textsc{Mrgp} & \cite{arnaldo2014multiple} (2014) & It decouples and
				linearly combines a program's subexpressions via multiple regression on the target variable  \\
				Local optimization NLS    & \cite{kommenda2018local} (2018)   & Constants Optimization in GP by Nonlinear Least Square \\
				\textsc{Feat} & \cite{la2018learning} (2018) & Features are represented as networks of multi-type expression trees  comprised of activation functions; Differentiable features are trained via gradient descent \\
				\textsc{Sbp-Gp} &  \cite{virgolin2019linear} (2019)  & The idea of semantic back-propagation utilized in GP \\
				\textsc{Bsr} & \cite{jin2019bayesian} (2019) & ML--based approach; Bayesian symbolic regression  \\ 
				\textsc{Dsr}  & \cite{petersen2019deep} (2019) & Deep Symbolic Regression based on a RNN approach further utilizing the policy gradient search \\
				\textsc{Operon} & \cite{kommenda2020parameter} (2020) &  Utilizing nonlinear least squares for parameter identification of SR models with LS \\
				\textsc{Vnp}    & \cite{elleuch2020variable} (2020) & VNS based GP approach \\
				\textsc{OccamNet} & \cite{costa2020fast} (2020) &   A fast neural network approach; the model defines a probability distribution over a non-differentiable function space; it samples functions and updates the weights with back-propagation  based on cross-entropy matching in an EA strategy	 \\

				\textsc{AI-Feynman} & \cite{udrescu2020ai} (2020) & A physics-inspired divide-and-conquer method; it  combines neural network fitting \\
				\textsc{Gomea}  & \cite{virgolin2021improving} (2021)    & A model-based
				EA framework called Gene-pool optimal mixing evolutionary algorithm \\
				\textsc{Itea} & \cite{de2021interaction} (2021)   & EA based approach called the interaction-transformation EA   \\
				\textsc{Sa} & \cite{kantor2021simulated} (2021) &  Simulated annealing approach \\

				\textsc{Drla} & \cite{mundhenk2021symbolic} (2021)  &    A deep reinforcement learning approach enhanced with genetic programming \\
				\textsc{Taylor-Gp} &  \cite{he2022taylor} (2022)  &  Taylor polynomials approximations  \\ \hline

		\end{tabularx} }
		\caption{Overview of methods to solve SR.}
		\label{tab:gp-based}
	\end{table}

   The progress in the SR field suffered from a lack of uniform, robust, and transparent
	benchmarking standards over the last two decades. Recently, La Cava et al.~\cite{la2021contemporary} proposed an
	open-source, reproducible benchmarking platform for SR, called SRBench. The authors extended  PMLB~\cite{olson2017pmlb},  a repository of standardized regression problems, by 130 SR datasets for which model forms are known; see more in the aforementioned paper. In the extensive experimental evaluation  where 14
	symbolic regression methods and 7 machine learning methods are compared on the set of 252 diverse regression problems. The general conclusions derived from there are: ($i$) when comes to real-world problems, the best performing algorithms are those which combine genetic algorithm with a parameter estimation; ($ii$) when comes to the instances with a presence of noise, the GP--based and deep learning techniques are performing similarly.

	In this work we present a novel approach to solve SR, which combines the popular iterated local search meta-heuristic (ILS)~\cite{lourencco2003iterated,lourencco2019iterated} with the ordinary least square  method (OLS)~\cite{leng2007ordinary}. Briefly, ILS handles combinatorial (discrete) aspects of search space, while OLS helps in the process of a coefficient determination, so it handles continuous parts of the search space. As will be shown later, the proposed method has shown its robustness w.r.t. introduced noise, therefore we called the method \textsc{Rils}-\textsc{Rols}  (with letters R corresponding to regression and robust).    Additionally, in order to navigate the search towards not only accurate models, but increasing the chances to get the exact one, the algorithms is equipped with a carefully constructed fitness function that utilizes three important scores of a solution (model): RMSE and $R^2$ scores which participate into models accuracy, and a weighted solution size that will contribute penalizing large (mostly non-exact) solutions.

	The main contributions of this work may be summarized as follows:

	\begin{enumerate}
		\item The proposed method outperforms 14 state-of-the-art methods for the problem instances from the two ground-truth benchmark sets from literature, used in the benchmarking platform SRBench.

		\item The method shows its high robustness, which is proved by its comparison to the other algorithms in presence of different levels of Gaussian white noise in the input data.

		\item The method is very efficient, taking in average less than 3 minutes to reach the exact solution, on the problem instances where it was possible. For example, the second best approach in terms of the delivered exact percentage, \textsc{AI-Feynman}, needs significantly more time, which is around 45 minutes.

		\item A new set of unbiased instances is introduced that consists of randomly generated formulae of various sizes and number of input variables presenting exact models that need to be found. This set was employed to analyze the effect of the model size, the number of input variables in the model and the level of noise on the solving difficulty. 
	\end{enumerate}

@@

	Before we give details on our \textsc{Rils}-\textsc{Rols}  method for solving SR, we will explain the basics of two main techniques incorporated into our algorithm. These are ILS meta-heuristic and OLS methods.

@@

	The most popular version of ILS is shown in Algorithm~\ref{alg:ils}. Note that we use this version of ILS as the backbone of our \textsc{Rils}-\textsc{Rols}  algorithm.

@@

	Thus, we get the system of equations: 
	\begin{align*}
		% &\sum -2X_i (\tilde{y}_i - y_i )  = 0 \\
		& -2X_1 (y_1 - \tilde{y}_1)  = 0 \\
		& \vdots \\
		& -2X_d (y_d - \tilde{y}_d)  = 0 \\
		&\sum -2  (y_i - \tilde{y}_i) = 0.
	\end{align*}
	%The above system can be transformed to 
	%\begin{align*}
	%	 & \sum y_i \cdot x_i   = b\sum x_i \cdot x_i  +  k \cdot \sum x_i \\
	%	 & \sum y_i  = k \sum x_i + b \cdot d. 
	%  \end{align*}
By solving the above system of $(d+1)$ equations, one can obtain vector of coefficients $k=(k_1, ..., k_d)$ and intercept $b$. %For example, in order to calculate coefficient $k_i$, the following system of equations is 
%\fxnote{TODO: Aca -- ovde je pisalo da je to sistem sa dve jednacine i dve nepoznate k i b -- medjutim, k je vektor. Ako mozes dopisi jos nesto, tipa koliko je podataka potrebno i slicno -- mozda veremenska slozenost.) }%https://towardsdatascience.com/understanding-the-ols-method-for-simple-linear-regression-e0a4e8f692cc
\subsection{\textsc{Rils}-\textsc{Rols}  method}
Now we will explain the proposed \textsc{Rils}-\textsc{Rols}  method in detail. The overall method scheme is given in Algorithm~\ref{alg:rilsrols}.

>>>>

	% first the title is needed
	\title{\textsc{RILS}-\textsc{ROLS} : Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares}

	\author[1]{Aleksandar Kartelj}
	\author[2]{Marko Djukanovi\'c}
	\address[1]{$kartelj@matf.bg.ac.rs$, \\  Faculty of Mathematics, University of Belgrade, Serbia}
	\address[2]{$ marko.djukanovic@pmf.unibl.org$,\\   Faculty of Natural Sciences and Mathematics, University of Banja Luka, Bosnia and Herzegovina}
	\begin{abstract}
		In this paper, we solve the well-known symbolic regression problem. This problem has been intensively studied in the literature and has a wide range of applications. The symbolic regression models could provide insights towards establishing physical theory behind experimentally observed data. Practical applications include revealing complex ecological dynamics in ecology, discovering mutations effects on protein stability in biology, modeling analytic representations of the exciton binding energy in energy science, finding models for band gaps of NaCl-type compounds
		in material science, just to name a few.

		We propose a new meta-heuristic-based approach, called \textsc{RILS}-\textsc{ROLS} to solve this problem. \textsc{RILS}-\textsc{ROLS} is based on the following elements: ($i$) iterated local search is the method backbone, mainly solving combinatorial and some continuous aspects of the problem; ($ii$) ordinary least square method is focused on the continuous aspect of the search space -- it efficiently determines the best--fitting coefficients of linear combinations within solution equations;  ($iii$) a novel fitness function combines three important model quality measures: $R^2$ score, RMSE score, and the size of the model (or model complexity).

		The experiments are conducted on the two well-known ground-truth benchmark sets called \textsc{Feynman} and \textsc{Strogatz}. \textsc{RILS}-\textsc{ROLS} was compared to 14 other competitors from the literature with  presence of different levels of the Gaussian white noise in the input data. Our method outperformed all 14 competitors with respect to the true symbolic model accuracy percentage under all level of noise. 
		In addition to evaluation on known ground-truth datasets, a new randomly generated set of problem instances is introduced. The goal of \textsc{Random} dataset was to test the scalability of our method with respect to incremental equation sizes and different levels of noise. \textsc{RILS}-\textsc{ROLS} can be considered as a new state-of-the-art method for solving the problem of symbolic regression when ground-truth equations are known.
	\end{abstract}
	\maketitle

@@

	The problem of symbolic regression (SR)~\cite{billard2002symbolic} has attracted many researchers over the last decade. SR can be seen as a generalization of more specific variants of regression in which the functional form is fixed, e.g. the well-known linear regression, polynomial regression~\cite{stimson1978interpreting}, etc. All regression models have the same goal: given a set of $n$-dimensional input data and its corresponding continuous output target variable, the aim is to find a  mathematical expression (function) of $n$ (input) variables that best \emph{fits} the target variable.  %This computationally intensive task is in general provenly NP--hard~\cite{virgolin2022symbolic}. 
	In the case of linear regression, the model is always a linear combination of input variables. This is in general not good enough, since target variable might be dependent on some nonlinear function among input variables. Unlike linear regression, SR allows the search over much larger space of possible mathematical formulas to find the best-fitting ones, i.e. those able to predict the target variable from the input variables. The basis of constructing explicit formula is in elementary operations like addition and multiplication, as well as polynomial, trigonometric, exponential, and others.  
	%Application: https://towardsdatascience.com/real-world-applications-of-symbolic-regression-2025d17b88ef

	SR models tend to be interpretable, unlike artificial neural networks. Coefficients inside SR formulas might indicate the absolute or relative importance of certain input variables. Moreover, relations among variables might be described in an elegant way. For example, the appearance of an exponential function can be associated with some physical phenomenon such as the intensity of radiation or acceleration over time, see~\cite{udrescu2020ai}. Additionally, the SR models often have high generalization power unlike some models with fixed functional forms such as polynomial regression.

	Practical applications of SR in chemical and biological sciences are listed in~\cite{weng2020simple}. In particular, this paper describes the discovery of a series of new oxide perovskite catalysts with improved activities. The application of SR to discovering physical laws from a distorted video is studied in~\cite{udrescu2021symbolic}. Revealing complex ecological dynamics by SR is presented in~\cite{chen2019revealing}. Paper \cite{louis2021reviewing} shows how SR is used to model the effects of the mutations on protein stability, in the domain of fundamental and applied biology. One recent study \cite{liu2021machine} is concerned of auto-discovering conserved quantities using trajectory data from unknown dynamical systems. Paper~\cite{liang2019phillips} shows the application of SR to model analytic representations of the exciton binding energy. The use of SR in material science is described in~\cite{wang2019symbolic,wang2022symbolic,burlacu2022symbolic,kabliman2021application}. SR application to wind speed forecasting is given in~\cite{abdellaoui2021symbolic}.

	There are many different ways to tackle the SR. Most of them are based on the machine learning techniques, genetic programming (GP) or some other meta-heuristics. Among the first GP methods to tackle SR is the one of Raidl~\cite{raidl1998hybrid} based on a hybrid variant of genetic programming. A differential evolution algorithm was proposed by Cerny et al.~\cite{cerny2008using}. Age-fitness Pareto Optimization approach is proposed by Smidt and Lipson~\cite{schmidt2010age}.  Application of the artificial bee colony programming to solve SR is proposed by Karaboga et al.~\cite{karaboga2012artificial}. 
	Application of local-based heuristics for solving SR is reported by Commenda in his PhD thesis~\cite{kommenda2018local}. A GP-based approach, the gene-pool optimal mixing evolutionary algorithm (\textsc{GOMEA}) is studied by Virgolin et al.~\cite{virgolin2021improving}.  Another evolutionary algorithm, the interaction-transformation EA (\textsc{ITEA}) has been proposed by de Franca et al.~\cite{de2021interaction}. Simulated annealing to solve SR is proposed by Kantor~\cite{kantor2021simulated}. A variable neighbourhood programming approach to solve SR  is proposed by Elleurich et al.~\cite{elleuch2020variable}; this technique is initially proposed in~\cite{elleuch2016variable}. 
	Kommenda et al.~\cite{kommenda2020parameter} proposed a method called \textsc{Operon} algorithm, which uses nonlinear least squares for parameter identification of SR models further integrated into a local search mechanism in tree-based GP. The C++ implementation of \textsc{Operon} is discussed in~\cite{burlacu2020operon}. The method that utilizes the Taylor polynomials to approximate the symbolic equation that fits the dataset, called Taylor genetic programming, is proposed in~\cite{he2022taylor}. A GP approach that uses the idea of semantic back-propagation (Sbp-Gp) is proposed in~\cite{virgolin2019linear}.  Empirical analysis of variance between many GP-based methods for SR is discussed in~\cite{kammerer2021empirical}. It is also worth to mention the GP-based Eurequa commercial solver~\cite{schmidt2009distilling, schmidt2011machine} that uses age-fitness pareto optimization with co-evolved fitness estimation. This solver is nowadays accepted as the gold standard of symbolic regression.

	Method based on Bayesian symbolic regression (\textsc{BSR}) is proposed in~\cite{jin2019bayesian} -- this method belongs to a family of Markov Chain Monte Carlo algorithms (MCMC). Deep Symbolic Regression (\textsc{DSR}), an RNN approach, is proposed in~\cite{petersen2019deep} -- it utilizes the policy gradient search. This mechanism of search is further investigated in~\cite{landajuela2021improving}. A fast neural network approach, called \textsc{OccamNet},  is proposed in~\cite{costa2020fast}.  A deep reinforcement learning approach enhanced with genetic programming is proposed in~\cite{mundhenk2021symbolic}.

    Powerful hybrid techniques to solve SR are also well studied; among them, we emphasize the \textsc{Eplex} solver from~\cite{la2019probabilistic,la2016epsilon} and \textsc{AI Feynman} algorithm from~\cite{udrescu2020ai}, a physics-inspired divide-and-conquer method combined with the neural network fitting. The latter is one of the most efficient methods for physically-inspired models. We also mention the Fast Function Extraction (\textsc{FFX}) algorithm developed by McConaghy~\cite{mcconaghy2011ffx}, which is a non-evolutionary method combined with a machine learning technique called path-wise-regularized learning, which quickly prunes a huge set of candidate basis functions down to compact models.

	A short overview of the most important literature methods to solve SR is given in Table~\ref{tab:gp-based}.

	\begin{table}[!h]
		\centering
		\scalebox{0.7}{
			\begin{tabularx}{550pt}{l  l  X}  
				Algorithm          &   Paper/year &   Short details   \\ \hline
				\textsc{GP}        &    \cite{koza1994genetic} (1994)    & Application of GP to SR \\ 
				\textsc{Hybrid-GP}  &   \cite{raidl1998hybrid} (1998)   & GP-based method; solutions are locally optimized with OLS to find optimum coefficients for top-level terms \\ 
				\textsc{DFE}        & \cite{cerny2008using} (2008)   & Differential evolution algorithm \\
				\textsc{APF-FE}                & \cite{schmidt2009distilling, schmidt2011machine} (2009, 2011) & Age-fitness Pareto optimization approach using co-evolved fitness estimation   \\
				APF            &      \cite{schmidt2010age} (2010)        &  Age-fitness Pareto optimization approach                        \\
				\textsc{FFX}  & \cite{mcconaghy2011ffx} (2011)  & The fast function extraction algorithm --  non-evolutionary technique based on a machine learning technique called path-wise regularized learning \\
				\textsc{ABCP} &  \cite{karaboga2012artificial} (2012)  & Artificial bee colony programming  approach \\
				\textsc{EPLEX}          &      \cite{la2016epsilon} (2016)         &   A parent selection method called $\epsilon$-lexicase selection  \\ 
				\textsc{MRGP} & \cite{arnaldo2014multiple} (2014) & It decouples and
				linearly combines a program's subexpressions via multiple regression on the target variable  \\
				Local optimization NLS    & \cite{kommenda2018local} (2018)   & Constants optimization in GP by nonlinear least squares \\
				\textsc{FEAT} & \cite{la2018learning} (2018) & Features are represented as networks of multi-type expression trees comprised of activation functions; differentiable features are trained via gradient descent \\
				\textsc{SBP-GP} &  \cite{virgolin2019linear} (2019)  & The idea of semantic back-propagation utilized in GP \\
				\textsc{BSR} & \cite{jin2019bayesian} (2019) & ML-based approach; Bayesian symbolic regression  \\ 
				\textsc{DSR}  & \cite{petersen2019deep} (2019) & Deep symbolic regression based on a RNN approach further utilizing the policy gradient search \\
				\textsc{Operon} & \cite{kommenda2020parameter} (2020) &  Utilizing nonlinear least squares for parameter identification of SR models with LS \\
				\textsc{VNP}    & \cite{elleuch2020variable} (2020) & VNS based GP approach \\
				\textsc{OccamNet} & \cite{costa2020fast} (2020) &   A fast neural network approach; the model defines a probability distribution over a non-differentiable function space; it samples functions and updates the weights with back-propagation  based on cross-entropy matching in an EA strategy	 \\

				\textsc{AI-Feynman} & \cite{udrescu2020ai} (2020) & A physics-inspired divide-and-conquer method; it also uses neural network fitting \\
				\textsc{GOMEA}  & \cite{virgolin2021improving} (2021)    & A model-based
				EA framework called gene-pool optimal mixing evolutionary algorithm \\
				\textsc{ITEA} & \cite{de2021interaction} (2021)   & EA based approach called the interaction-transformation EA   \\
				\textsc{SA} & \cite{kantor2021simulated} (2021) &  Simulated annealing approach \\

				\textsc{DRLA} & \cite{mundhenk2021symbolic} (2021)  &    A deep reinforcement learning approach enhanced with genetic programming \\
				\textsc{Taylor-GP} &  \cite{he2022taylor} (2022)  &  Taylor polynomials approximations  \\ \hline

		\end{tabularx} }
		\caption{SR methods overview.}
		\label{tab:gp-based}
	\end{table}

   The SR researches lack uniform, robust, and transparent
	benchmarking standards. Recently, La Cava et al.~\cite{la2021contemporary} proposed an
	open-source, reproducible benchmarking platform for SR, called SRBench. The authors extended  PMLB~\cite{olson2017pmlb},  a repository of standardized regression problems, with 130 SR datasets for which exact (\emph{ground-truth}) models are known; see more in the aforementioned paper. In their extensive experimental evaluation, 14
	symbolic regression methods and 7 machine learning methods are compared on the set of 252 diverse regression problems. The remaining 122 SR datasets belong to a class of so-called \emph{black-box} problems for which the \emph{ground-truth} is unknown. One of the most interesting conclusions from La Cava et al. research is that algorithms specialize in either solving \emph{ground-truth} problems, or \emph{black-box} problems, but not both.

	In this work we present a novel approach to solve \emph{ground-truth}-based SR, which combines the popular iterated local search meta-heuristic (ILS)~\cite{lourencco2003iterated,lourencco2019iterated} with the ordinary least square method (OLS)~\cite{leng2007ordinary}. ILS mostly handles combinatorial (discrete) aspects of search space, while OLS helps in the process of a coefficient determination, so it handles some continuous parts of the search space. As will be shown later, the proposed method has shown its robustness w.r.t. introduced noise, therefore we called the method \textsc{RILS}-\textsc{ROLS}  (with letters R corresponding to regression and robust). Additionally, in order to navigate the search toward exact model (and not only accurate one) the algorithm is equipped with a carefully constructed fitness function that utilizes three important model characteristics: RMSE and $R^2$ scores which participate into models accuracy, and a weighted solution size that penalizies too complex models.

	The summary of main contributions is as follows:

	\begin{enumerate}
		\item The proposed method outperforms 14 comparison methods on two \emph{ground-truth} benchmark sets from literature -- these methods and benchmarks are used in SRBench platform.

		\item The method shows its high robustness, which is proved by its comparison to the other algorithms in the presence of different levels of Gaussian white noise in the input data.

		\item The method is very efficient, taking in average less than 4 minutes to reach the exact solution, on the problem instances where it was possible.

		\item A new set of unbiased instances is introduced -- it consists of randomly generated formulae of various sizes and number of input variables. This set was employed to analyze the effect of the model size and the level of noise on the solving difficulty. 
	\end{enumerate}

@@

	Before we give details on our \textsc{RILS}-\textsc{ROLS}  method for solving SR, we will explain the basics of two main techniques incorporated into our algorithm. These are ILS meta-heuristic and OLS methods.

@@

	The most popular version of ILS is shown in Algorithm~\ref{alg:ils}. Note that we use this version of ILS as the backbone of our \textsc{RILS}-\textsc{ROLS}  algorithm.

@@

	Thus, we get the system of equations: 
	\begin{align*}
		% &\sum -2X_i (\tilde{y}_i - y_i )  = 0 \\
		& -2X_1 (y_1 - \tilde{y}_1)  = 0 \\
		& \vdots \\
		& -2X_d (y_d - \tilde{y}_d)  = 0 \\
		&\sum -2  (y_i - \tilde{y}_i) = 0.
	\end{align*}
	%The above system can be transformed to 
	%\begin{align*}
	%	 & \sum y_i \cdot x_i   = b\sum x_i \cdot x_i  +  k \cdot \sum x_i \\
	%	 & \sum y_i  = k \sum x_i + b \cdot d. 
	%  \end{align*}
By solving the above system of $(d+1)$ equations, one can obtain vector of coefficients $k=(k_1, ..., k_d)$ and intercept $b$. %For example, in order to calculate coefficient $k_i$, the following system of equations is 
%\fxnote{TODO: Aca -- ovde je pisalo da je to sistem sa dve jednacine i dve nepoznate k i b -- medjutim, k je vektor. Ako mozes dopisi jos nesto, tipa koliko je podataka potrebno i slicno -- mozda veremenska slozenost.) }%https://towardsdatascience.com/understanding-the-ols-method-for-simple-linear-regression-e0a4e8f692cc
\subsection{\textsc{RILS}-\textsc{ROLS}  method}
Now we will explain the proposed \textsc{RILS}-\textsc{ROLS}  method in detail. The overall method scheme is given in Algorithm~\ref{alg:rilsrols}.
